Detecting skill level and mental templates in Late Acheulean biface morphology: Archaeological and experimental insights

### 2022-07-24 manuscript/manuscript.Rmd
revised the introduction and added the correct figures.

---
title: "Detecting skill level and mental templates in Late Acheulean biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental templates have been frequently cited. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving twenty-six naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern expert requires more training time than was permitted in this extensive and long-running training program. Our study demonstrated the potential of experimental archaeology and digital photographs in revealing new insights from old archaeological assemblages. \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021]. From this extensive list, skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.

In its classical definition, mental template is the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept essentially lies in the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of bifaces at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraints [@white1995], and the lack of rigorous studies of its corresponding cognitive processes. Until very recently, several researchers have actively addressed these critiques and reconceptualized the mental template in the study of biface morphology. Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Furthermore, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental template. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in archaeological context in composition, size, and shape. In terms of the cognitive mechanisms behind mental template, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations, which has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors. In short, when exercised with proper caution, mental template still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness and elongation, among other things.

Building upon the concept of mental template, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners. It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995]. When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of biface technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers. Since the novice biface collection is generated from a 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental template be efficiently detected from biface morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples prepared by R. I. [@iovita2011], which include both the front and lateral views of a given specimen. First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting. This is then followed by the batch conversion of color photographs to a black-and-white binary format. Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop. The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom ImageJ script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette. The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements). Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006], which also at least alleviate the effect of resharpening process to some extent. The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al. [-@pargeter2019].

@@

@@

Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R. Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas. Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data reuse is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].

@@

**Cheng Liu**: Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing. **Nada Khreisheh**: Investigation, Writing -- review & editing. **Radu Iovita**: Resources, Writing -- review & editing. **Dietrich Stout**: Conceptualization, Investigation, Resources, Funding acquisition, Supervision, Writing -- original draft, Writing -- review & editing. **Justin Pargeter**: Conceptualization, Investigation, Methodology, Supervision, Writing -- original draft, Writing -- review & editing.

@@

This work was supported by funding from the National Science Foundation of the USA (grants SMA-1328567 & DRL-1631563), the John Templeton Foundation (grant 47994), and the Emory University Research Council. The handaxe knapping skill acquisition experiment involved in this study was approved by Emory University's Internal Review Board (IRB study no: 00067237).

>>>>

---
title: "Detecting skill level and mental templates in Late Acheulean biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental templates have been frequently cited. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving twenty-six naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. Our study demonstrated the potential of experimental archaeology and digital photographs in revealing new insights from old archaeological assemblages. \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021]. From this extensive list, skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of bifaces at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes. To avoid the historical baggage associated with this controversial term, some researchers have also developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

Until very recently, several researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of biface morphology. Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning. Furthermore, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in archaeological context in composition, size, and shape. In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources., Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains. In short, when exercised with proper caution, the concept of mental template still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness and elongation, among other things.

Following the reconceptualization of the concept of mental template, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners. It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995]. When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of biface technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers. Since the novice biface collection is generated from a 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental template be efficiently detected from biface morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples [@iovita2011], which include both the front and lateral views of a given specimen. First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting. This is then followed by the batch conversion of color photographs to a black-and-white binary format. Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop. The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom ImageJ script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette. The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements). Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006], which also at least alleviate the effect of resharpening process to some extent. The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al. [-@pargeter2019].

@@

In addition, visual inspection of the principle component scatter plot (Fig. ) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups. To test this,  we  conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)). Across all three groups, a negative correlation has been displayed between the PC1 and PC2 values, although this trend is not statistically significant (r=-0.41, p= 0.24) in the expert group, probably because of its small sample size.

```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132)."}
knitr::include_graphics(here::here("figure/PC correlation.png"))
```

@@

Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas. Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data reuse is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].

@@

**Cheng Liu**: Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing. **Nada Khreisheh**: Investigation, Writing -- review & editing. **Dietrich Stout**: Conceptualization, Investigation, Resources, Funding acquisition, Supervision, Writing -- original draft, Writing -- review & editing. **Justin Pargeter**: Conceptualization, Investigation, Methodology, Supervision, Writing -- original draft, Writing -- review & editing.

@@

This work was supported by funding from the National Science Foundation of the USA (grants SMA-1328567 & DRL-1631563), the John Templeton Foundation (grant 47994), and the Emory University Research Council. The handaxe knapping skill acquisition experiment involved in this study was approved by Emory University's Internal Review Board (IRB study no: 00067237). We would also like to thank Radu Iovita for providing us access to the digital photographs of the Boxgrove handaxe assemblage.
